Predicting the effects of character displacement on food-web dynamics

### 2018-08-12 ECD_Supp_Mat.Rmd
adding intro from some old written text.

>>>>

Keywords: competition; eco-evolutionary dynamics; ecological speciation; three-spine stickleback; zooplankton

## Abstract

Ecological character displacement is a key driver of diversification and adaptive radiation; however, there has been little work examining the consequences of character displacement on the dynamics of ecological communities. Given the growing recognition for the role of evolution in structuring ecological communities, understanding how major evolutionary processes impact ecological dynamics is increasingly important. Here, we build a mathematical model to predict how exploitative competition between two consumers for two resources affects the evolution of consumer attack rates, and, in turn, consumer-resource population dynamics. We then compared these predictions to results from a natural system, where a fish consumer (three-spine stickleback) has become locally adapted to different aquatic habitats with different resources (benthic vs. limnetic zones). Our model suggested that increasing ecological character displacement decreases the equilibrium density of resources, and with sufficient divergence, results in more oscillatory (destabilizing) consumer-resource dynamics. In support of this prediction, we found that the biomass density of zooplankton (resource) in lakes with two species of stickleback was only a fraction of that found in lakes with one species, varying from 4 to 41%. Moreover, we found that the biomass density of zooplankton was 2-fold more variable (CV) in lakes with two vs. one species of stickleback. These results suggest that ecological character displacement can eventually have a negative feedback on food-web persistence in the absence of other processes.

## Introduction

Ecological character displacement (ECD) is thought to be a key evolutionary process in the diversification of species and consequently the generation of biodiversity (Schluter 2000; Pfennig and Pfennig 2010; Stuart and Losos 2013). ECD describes the “process of phenotypic evolution in a species generated or maintained by [exploitative] resource competition with one or more coexisting species” (Schluter 2000). Over the past 40 years, a large body of theoretical (e.g. Lawlor and Smith 1976; Abrams 1986; Doebeli 1996; Taper and Chase 1985) and empirical (reviewed in: Schluter 2000; Dayan and Simberloff 2005; Stuart and Losos 2013) work has been generated to understand the scenarios under which exploitative competition for resources leads to the divergence of consumer foraging traits. Yet, although the process of ecological character displacement is driven by the interplay between competition and trait change, and hence is strongly dependent on resource availability, little of the work to date has explicitly considered how character displacement impacts other species within the community. Similarly, research on food-web dynamics has largely ignored the potential role of evolution in altering interactions and stability (Fussmann et al. 2007). Given the emerging evidence that evolutionary processes can drive ecological dynamics (Hairston et al. 2005, Post et al. 2008, Harmon et al. 2009, Schoener 2011, Rudman et al. 2015), understanding the ecological consequences of character displacement is a natural first step in investigating how evolutionary processes affect food-web dynamics.

Ecological character displacement has typically been considered to be a stabilizing factor in communities (Lawlor and Smith 1976). This is because divergence in foraging traits is thought to reduce the strength of interspecific competition between consumers. Reducing the strength of interspecific competition relative to intraspecific competition is a stabilizing force in consumer-resource systems. Consumer specialization on distinct resources suggests that there are niche differences between consumers. Mathematically, this is also often assumed to be the case, because using techniques such as evolutionary invasion analysis describe the ability for a consumer with a mutant allele to increase in abundance when rare, a demographic signature of niche differentiation. Therefore, if both consumers are able to invade when rare, then this suggests that they can coexist. On the other hand, relatively specialized consumers have also been shown to be a strongly stabilizing force in food webs (McCann et al. 2005). This is because these more specialized consumers may generate asynchrony in the dynamics of patchily distributed resources. This asynchrony in resources then permits consumers to move to habitat patches with higher resource densities after they have depleted the resources in a different patch. If, for example, the consumers were not specialized, then the resources would exhibit the same dynamics, so when the resources at low densities, consumers would be unable to switch to alternative resources, which would cause the consumer populations to crash, and oscillate more with their resources (McCann et al. 2005). Consequently, whether ecological character displacement will tend to stabilize (via asynchronous resource dynamics) or destabilize (via increased interaction strength) food webs is unclear.

On the other hand, ecological character displacement often results in the evolution of characters that enhance a consumer’s ability to capture resources, which tends to destabilize consumer-resource interactions (Murdoch et al. 2003; McCann 2011).

This occurs because increases in a consumer’s attack rate results in the resource being suppressed well below its carrying capacity, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics (Rosenzweig 1971; Murdoch et al. 2003). Experimental decreases in interaction strength between consumers and resources has been shown to promote coexistence between consumers and resources (Luckinbill 1973). Moreover, there is broad support that specialized consumers, since they are tightly coupled to their resources, tend to exhibit consumer-resource cycles, whereas this is not the case for generalists (Murdoch et al. 2002). Therefore, although evolutionary responses to exploitative competition may reduce the strength of interactions between consumers, they ultimately increase the strength of interactions with their resources which can destabilize the direct consumer-resource interaction (rather than the indirect consumer-consumer interaction). While prior models of character displacement have included more realistic functional responses and spatial relationships between consumers and resources, they have focused more on whether these factors alter patterns of character displacement rather than the effects of character displacement on the stability of the ecological community. As a consequence, it is currently unclear whether ecological character displacement has a net stabilizing or destabilizing effect on ecological communities.

Here, we address this knowledge gap by building a mathematical model to examine how ecological character displacement affects consumer-resource dynamics. We then compared the predictions from our model to field data from one of the most well-studied examples of ecological character displacement in a natural system, threespine stickleback in lakes of southern British Columbia. We specifically sought to address two questions: (1) How does exploitative competition between two consumers for two resources affect the evolution of consumer attack rates? (2) How does the evolution of consumer attack rates affect community stability?
